Estimating the frequency of multiplets in single-cell RNA sequencing from cell-mixing experiments

### 2018-08-04 peerj_rebuttal_letter.md

- Paragraph starting on line 37. Another strong argument for using unequal proportions of cell types when assessing the accuracy of a scRNA-seq technique is that it is more representative of what you’d find inside of an organism. When working with blood or a tissue containing multiple cell types, it’s unlikely that those cell types will exists in equal proportions. Make sure to emphasize this at some point.

*Response:
This is a good point.
I have added a sentence explaining that in naturally occurring mixtures of cells of different types (such as blood or tissue), the different cell types will usually be present at unequal proportions.*

- At some point in the Introduction is important to make it clear that other people have intentionally used unequal proportions of cell types in QC analysis of an scRNA-seq technique. For example, see Figure 2F of [Rosenberg et al, 2017](https://www.biorxiv.org/content/biorxiv/early/2017/02/02/105163.full.pdf). However, your work is novel in that it provides a detailed explanation of how to calculate the multiplet frequency in these scenarios.

*Response:
This is a good point.
I have added a reference to the Rosenberg et al study in the Introduciton.*

@@

- Lines 44-47. Give some brief detail about the Poisson distribution for readers who may not be familiar with it.

*Response:
I have added text explaining that the Poisson distribution is accurate when cells are loaded into droplets randomly and independently.*

- A figure with diagrams showing how the scRNA-seq techniques work may be helpful to some readers.

*Response:
Given the wealth of different scRNA-seq techniques that can lead to Poisson loading, I have decided against trying to illustrate them graphically.
I agree that such diagrams are useful, but they are present in most of the references cited in the first paragraph of the Introduction.
My expectation is that this paper will be of interest primarily to experts who are already familiar with the basics of scRNA-seq and/or have read some of the references that explain these methods.*

@@

- Instead of calling R code from within Python (which only works for Python 2 and not Python 3, as far as I know) it would be best to just provide two scripts for readers to choose from that each contain the whole analysis pipeline with one script written purely in Python (no R code) and another purely written in R.

*Response:
This is a good idea. 
I now provide two separate Juypter notebooks, one that performs the calculations purely in R, and one that performs the calculations purely in Python.*

@@

### Experimental design
- See my previous comment regarding the Rosenberg et al paper.

*Response:
This point has been addressed; see above.*

- How closely do you expect the number of cells per droplet to follow a Poisson distribution? From my understanding of microfluidics devices, there would be some upper limit to how many cells could physically fit into a single droplet. This concern may be negligible statistically, but it should still be addressed.

*Response:
This is an excellent point. 
I am not able to directly quantify how "Poisson" different methods actually are from available data, so I have added several sentences to the Discussion that discuss how factors such as cell clumping could cause deviations from Poisson behavior, and how this might affect the calculations.*

@@

*Response: 
Thanks for the nice summary.*

### Comments for the author
I have a few questions and comments regarding the potential for more general application of the analysis described here:

1)  In many single-cell RNA-Seq experiments, unsupervised clustering reveals very discrete cell types that are readily distinguishable, but from the same species. The author focuses on the case that two cells from two different species are mixed in arbitrary proportion, but could these results be extended to the case that an arbitrary number of cell types is mixed in varying proportions? As a concrete example, if I am looking at a blood sample, I might expect to readily separate monocytes, B cells, T cells, etc. Could the framework described here be used to calculate an expectation value of, for example, the doublet frequency of each pair of cell types? This may be beyond the scope of the study described here, but may be worth a comment in the manuscript.

*Response:
This is an interesting idea.
My concern would be that in the case of cell-type clustering, the multiplets might affect the clustering and subsequently the cell-type counting.
In the case of mixing human and mouse cells, the classification is unambiguous in terms of single-species or mixed-species. 
But in cell type classification, if there are a substantial number of multiplets, they might skew the clustering, thereby changing the numbers of cells of each type that go into the calculation.
I am not sure how big a problem this would be in practice, but because of this concern I am reluctant to suggest the approach be applied in such a scenario without extensive further investigation.
For this reason, I have simply included sentences to the Discussion that emphasize that the method is restricted to cases where the cell types have already been classified.*

@@

2)  Throughout the paper, the author focuses on what I would call “statistical multiplets”. However, in many single-cell RNA-Seq experiments, this is not the only source of multiplets (and in some cases, not even the dominant source). For example, incomplete cellular dissociation could result in sequencing of multiplets simply because the cells were stuck together in the original tissue and remain stuck together during the profiling experiment. In my own work, I have noticed an enrichment in apparent multiplets coming from cell types that I know are interacting with each other in the tissue I am profiling. This makes the application described in 1) particularly important. Having an expectation value for the “statistical multiplet” frequency could provide a framework for evaluating whether or not the observed multiplet frequency for a pair of cell types is higher than one would expect, thereby implying incomplete dissociation. Again, this may be beyond the scope of the study described here, and I leave this to the author's judgement, but I think it would be worth commenting on the distinction between "statistical multiplets" and multiplets arising from other sources.

*Response:
This is an excellent point, and similar to one made by Reviewer 1 about the validity of the Poisson assumption.
I have added text to the Discussion (second paragraph) emphasizing how the calculation assumes a Poisson distribution which could be violated by factors such as cell clumping.
However, I am not certain how to extend the calculations to identify deviations from Poisson behavior with the input data that the approach currently assumes.
I think that an additional data point (such as the number of empty droplets) would need to be available in order to determine if the multiplet frequency is actually higher than expected under Poisson.
We can say that non-Poisson behavior should make the calculations here conservative provided that there is not a bias towards clumping of cells of the same type; this fact is now noted in the Discussion.*

### 2018-08-04 paper.tex
completed revision

The most common method to determine the multiplet frequency is to use a mix two types of cells (e.g., human and mouse).
During the analysis of the sequencing results, each non-empty partition can be identified as containing transcripts from one or both of the two cell types.
Partitions that contain a substantial number of transcripts from both cell types must be multiplets.
If the two cell types are mixed equally and the average number of cells per partition is low (so that most multiplets are doublets), then the multiplet frequency can be estimated as simply twice the fraction of non-empty partitions that contain a mix of cell types.
The logic is that all the multiplets are doublets, and only half the doublets will have cells of both types (the others will have two cells of the same type).
This approach has been used to estimate the multiplet frequency during the prototyping of most single-cell RNA sequencing methods~\citep{klein2015droplet,macosko2015highly,zheng2017massively,gierahn2017seq,cao2017comprehensive}.

However, in some cases the two cell types may be mixed in unequal proportions.
Unequal mixing could arise simply from error during cell counting, but it can also be a desirable aspect of experimental design.
For instance, if the researcher is actually interested in the human cells and simply wants to include an internal control to estimate the multiplet frequency during each new experiment, then (s)he may want to add fewer mouse cells so that most of the resulting data is for the human cells.
In addition, when analyzing naturally occurring mixtures of cells of multiple types, the different cell types will usually be present in unequal proportions.
But when the cells are mixed unequally, it is no longer valid to estimate the multiplet frequency as simply twice the fraction of non-empty partitions that contain a mix of both cell types.
Surprisingly, I could find no published descriptions of how to calculate the multiplet frequency from unequal mixes of two cell types.
Here I remedy this gap in the literature by deriving the equations to compute the multiplet frequency when the cells are mixed in arbitrary proportions under the assumption that the number of cells per partition is Poisson distributed.
This Poisson assumption is accurate when cells are loaded randomly and independently into partitions.

@@

\section*{Conclusions}
I have described how to calculate the multiplet frequency in single-cell RNA sequencing experiments in which two cell types are mixed in arbitrary proportions.
It is important to note that this calculation requires that the sequencing data have already been analyzed to determine whether each partition contains a non-negligible number of transcripts from each cell type, but many common analysis programs (such as the 10X \texttt{cellranger} pipeline) already do this.
It is also important to note that the approach in this paper only calculates the multiplet frequency---it does \emph{not} actually identify the multiplets so that they can be removed from downstream analyses.
For that purpose, other more sophisticated approaches have been developd~\citep{ilicic2016classification,stoeckius2017cell,kang2018multiplexed}.
Nonetheless, simply calculating the multiplet frequency from the data returned by standard pipelines such as the 10X \texttt{cellranger} is important for many purposes, and the results here enable that to be done regardless of the proportions at which the cell types are mixed.

>>>>

The most common method to determine the multiplet frequency is to mix two types of cells (e.g., human and mouse).
During the analysis of the sequencing results, each non-empty partition can be identified as containing transcripts from one or both of the two cell types.
Partitions that contain a substantial number of transcripts from both cell types must be multiplets.
If the two cell types are mixed equally and the average number of cells per partition is low (so that most multiplets are doublets), then the multiplet frequency can be estimated as simply twice the fraction of non-empty partitions that contain a mix of cell types.
The logic is that all the multiplets are doublets, and only half the doublets will have cells of both types (the others will have two cells of the same type).
This approach has been used to estimate the multiplet frequency during the prototyping of most single-cell RNA sequencing methods~\citep{klein2015droplet,macosko2015highly,zheng2017massively,gierahn2017seq,cao2017comprehensive}.

However, in some cases the two cell types may be mixed in unequal proportions.
Unequal mixing could arise simply from error during cell counting, or it could be an intentional aspect of experimental design~\citep{rosenberg2018single}.
For instance, if the researcher is actually interested in the human cells and simply wants to include an internal control to estimate the multiplet frequency during each new experiment, then (s)he may want to add fewer mouse cells so that most of the resulting data is for the human cells.
In addition, when analyzing naturally occurring mixtures of cells of multiple types, the different cell types will usually be present in unequal proportions.
But when the cells are mixed unequally, it is no longer valid to estimate the multiplet frequency as simply twice the fraction of non-empty partitions that contain a mix of both cell types.
Surprisingly, I could find no published descriptions of how to calculate the multiplet frequency from unequal mixes of two cell types.
Here I remedy this gap in the literature by deriving the equations to compute the multiplet frequency when the cells are mixed in arbitrary proportions under the assumption that the number of cells per partition is Poisson distributed.
This Poisson assumption is accurate when cells are loaded randomly and independently into partitions.

@@

\section*{Conclusions}
I have described how to calculate the multiplet frequency in single-cell RNA sequencing experiments in which two cell types are mixed in arbitrary proportions.
It is important to note that this calculation requires that the sequencing data have already been analyzed to determine whether each partition contains a non-negligible number of transcripts from each cell type, but many common analysis programs (such as the 10X \texttt{cellranger} pipeline) already do this.

The calculation also assumes that the number of cells per droplets follows a Poisson distribution.
While many single-cell RNA sequencing methods are designed to partition cells in a way that concords with this assumption~~\citep{klein2015droplet,macosko2015highly,zheng2017massively,gierahn2017seq,cao2017comprehensive} , it is possible that cell clumping or other factors could bias certain partitions to contain more cells than expected under a Poisson distribution.
In such a scenario, the calculations in this paper would overestimate the true multiplet frequency if the clumping is equally likely across cell types, but could underestimate the true multiplet frequency if intra-cell-type clumping is more likely than inter-cell-type clumping.

Finally, the approach in this paper only calculates the multiplet frequency---it does \emph{not} actually identify the multiplets so that they can be removed from downstream analyses.
For that purpose, other more sophisticated approaches have been developd~\citep{ilicic2016classification,stoeckius2017cell,kang2018multiplexed,wolock2018scrublet,depasquale2018doubletdecon}.
Nonetheless, simply calculating the multiplet frequency from the data returned by standard pipelines such as the 10X \texttt{cellranger} is important for many purposes, and the results here enable that to be done regardless of the proportions at which the cell types are mixed.
